Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-26 aistats.tex
Change fig 1 and minor fixes

% TODO
%\begin{figure}[t]
%	\centering
%\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
%	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
%		Solid curve  are average over $10^5$ trials.
%		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
%		Shaded areas are 90\% confidence interval.
%		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
%		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
%		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
%	}
%	\label{fig:curves}
%\end{figure}

\begin{figure}[t]
	\centering
\includegraphics[width=.4\textwidth]{fewsamples.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance MLE (\S\ref{ssec:gaussian-variance}) against number of samples $n$. The bold curve plots the average over 10000 trials,  dark shaded area is 90\% confidence interval, and light shade is a min-max interval.
		The expected value is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}, while the variance remains large.
		We wish to find similar upper bounds for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

\paragraph{Statistical decision theory}
To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
We thus consider the fundamental problem: %TODO thus repetition
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g., $X$ is Gaussian with known variance)
or close to quadratic (\S\ref{ssec:quadratic}).
However, a  general solution for finite $n$ remains elusive. %TODO add link to figure 1
In this paper, we review partial solutions and give ideas on how to solve the problem.

@@

\subsection{Full Gaussian (Non-Trivial)}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, consider the full Gaussian $\cN(m,\sigma^2)$, which offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\  % TODO remove factor 2
	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
\citep[p.177]{nesterov2003introductory} % TODO add citation to example 4.1.1.4
, but $\logpart$ is not (see App.~\ref{app:gaussian}).
We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.

@@

\begin{figure}[ht]
	\centering
\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
		Solid curve  are average over $10^5$ trials.
		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, tight up to a factor 2).
		Shaded areas are 90\% confidence interval.
		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
	}
	\label{fig:MAP_rate}
\end{figure}

In this section, we prove the results mentioned in \S\ref{ssec:gaussian-variance},
and add some context and experimental observations.
In particular, we show the bound~\eqref{eq:MAP_rate} in \cref{fig:MAP_rate}.
Let us recall the form of the divergence with different parametrizations.
For a Gaussian with known mean $0$ and covariance $\sigma^2$, 
natural and mean parameters $\theta, \m$;
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right)
	= \phi\paren{\frac{\sigma_*^2}{ \sigma_n^2}},
	\\
	\bregmanconj(\m_*; \m_n) 
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) 
	= \phi\paren{\frac{\m_*}{ \m_n}},
	\\
	\bregman( \nat_n; \nat_* ) 
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right)
	= \phi\paren{\frac{ \nat_n}{\nat_*}},
\end{align}
where all these lines are equal, and $\phi$ measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)),
\end{align}
illustrated in orange in Figure~\ref{fig:phi}.
To upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focus on the symmetrized Bregman instead
\alignn{
	\bregmanconj( \mu_*; \mu_n) 
	\leq \cS_{\conj} (\mu_*, \mu_n ) 
	:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) 
	= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
	\label{eq:symmetrized_bregman}
}
which can also be expressed with $\phi$ as follows:
\aligns{
	\label{eq:log_bound}
	\phi(z) \leq \phi(z) + \phi\paren{\inv{z}}
	&=  \half \paren{z -1  + \inv{z} -1} 
	= \frac{(z-1)^2}{z} \; .
}
See the grey upper bound in Figure~\ref{fig:phi}.

>>>>

\begin{figure}[t]
	\centering
\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
		Solid curve  are average over $10^5$ trials.
		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
		Shaded areas are 90\% confidence interval.
		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
	}
	\label{fig:curves}
\end{figure}

@@

\paragraph{Statistical decision theory}
To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research,
and raising the fundamental problem:
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}

There are already general asymptotic results (\S\ref{ssec:asymptote} and Fig.~\ref{fig:curves}), and a finite $n$ result when $\logpart$ is quadratic (e.g., $X$ is Gaussian with known variance)
or close to quadratic (\S\ref{ssec:quadratic}).
However, a  general solution for finite $n$ remains elusive. 
In this paper, we review partial solutions and give ideas on how to solve the problem.

@@

\subsection{Full Gaussian (Non-Trivial)}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, consider the full Gaussian $\cN(m,\sigma^2)$, which offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\ 
	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
\citep[p.177, example 4.1.1.4]{nesterov2003introductory}
, but $\logpart$ is not (see App.~\ref{app:gaussian}).
We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.

@@

In this section, we prove the results mentioned in \S\ref{ssec:gaussian-variance},
and add some context and experimental observations.
Let us recall the form of the divergence with different parametrizations.
For a Gaussian with known mean $0$ and covariance $\sigma^2$, 
natural and mean parameters $\theta, \m$;
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right)
	= \phi\paren{\frac{\sigma_*^2}{ \sigma_n^2}},
	\\
	\bregmanconj(\m_*; \m_n) 
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) 
	= \phi\paren{\frac{\m_*}{ \m_n}},
	\\
	\bregman( \nat_n; \nat_* ) 
	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right)
	= \phi\paren{\frac{ \nat_n}{\nat_*}},
\end{align}
where all these lines are equal, and $\phi$ measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)),
\end{align}
illustrated in orange in Figure~\ref{fig:phi}.
To upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focus on the symmetrized Bregman instead
\alignn{
	\bregmanconj( \mu_*; \mu_n) 
	\leq \cS_{\conj} (\mu_*, \mu_n ) 
	:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) 
	= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
	\label{eq:symmetrized_bregman}
}
which can also be expressed with $\phi$ as follows:
\aligns{
	\label{eq:log_bound}
	\phi(z) \leq \phi(z) + \phi\paren{\inv{z}}
	&=  \half \paren{z -1  + \inv{z} -1} 
	= \frac{(z-1)^2}{z} \; .
}
See the grey upper bound in Figure~\ref{fig:phi}.
